---
title: "Broughton ecological classification"
author: "Edward Gregr"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    keep_tex: true       # Keep intermediate LaTeX file for troubleshooting
    latex_engine: pdflatex # Specify the LaTeX engine (pdflatex, xelatex, or lualatex). This renders the pdf
    number_sections: true # Optional, if you want numbered sections
    toc: true             # Table of contents (optional)
    fig_caption: true     # Enable figure captions
fontsize: 11pt            # Set font size (optional)
geometry: margin=1in      # Set margins (optional)
header-includes:
  - \usepackage{booktabs} # for tables
  - \usepackage{pdflscape}
  - \usepackage{tocloft}
---

```{r Introduction, include=FALSE, cache= FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 7, fig.height = 5.5,dpi = 300, out.width = '100%', fig.align="center")
knitr.table.format = 'html'
```
As part of the BC-funded climate ready seafood program, this work provides a regional classification of the Broughton region that will provide a summary of local physical data relevant to local communities. We hope it will provide a pathway to weave together scientific and Indigenous knowledge, while also underpinning the spatial analyses envisioned by the Local Seaweed Services Model (LSSM).

# Objective
This work extends the DFO classification project to a small, self-similar area within the Broughton Archipelago.

The objective is to classify the study area into a polygon representation suitable for supporting the spatial analyses of the LSSM. We will; 1) standardize available data sets to best reasonable spatial resolution (likely the local 20 x 20 m2 bathymetric grid - Davies et al. 2020); and 2) Prepare a classification analysis examining how spatial extents influence clusters created with the k-means clustering algorithm.


## Background   
Following the methods of Mora-Soto et al. 2024, create a k-means clustering of biophysical layers to support the representation of kelp habitat in British Columbia, with a particular focus on the Broughton region.

## Data summary

# Source data

Dimensions of the loaded Raster stack.
```{r echo=FALSE}
dim(prepped_layers)
```

```{r echo=FALSE, fig.cap="__*Figure 1: Maps of predictors scaled and trimmed to area of interest.*__"}
plot( prepped_layers )
par(mfrow = c(1, 1))
```

```{r echo=FALSE, fig.cap="__*Figure 2: Histograms of predictor data.*__"}
raster::hist(prepped_layers, nclass=50)
par(mfrow = c(1, 1))
```

Examine correlations across predictors, and identify those beyond a threshold of 0.6.
```{r echo=FALSE }

# Global: cor_table 

y_low <- lower.tri(cor_table, diag = TRUE)
cor_table[ y_low ] <- NA
knitr::kable( y, caption = "Correlation matrix for assessing predictor cross-correlations")

x <- 0.6
high_rows <- apply(cor_table, 1, function(row) any(row > x, na.rm = TRUE))
z <- cor_table[ high_rows, ]
knitr::kable( z, digits = 2, format = "latex", booktabs = TRUE, 
              caption = "Predictor variables that exceed 0.6 threshold") %>%
  kable_styling( latex_options = c("scale_down") ) %>% 
  row_spec(0, angle = 90) %>% landscape()

```

```{r Part2_ScreePlot, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="__*Figure 3: Scree plot showing the total within sum-of-squares across a range of cluster numbers.*__"}
plotme
```

```{r Part2_HeatMap, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="__*Figure 4: Heat map showing within cluster standard deviation of the predictors.*__"}
z_heat
```

```{r Part2_DataSummary, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="__*Figure 5: Silhouette plot showing pixel membership in each cluster and silhouette widths.*__"}
plot(sk, col = 1:nclust, border=NA )
```


# Results

## Part 1: Metrics across all prey 
### Metrics by month

__*Table 1a, b: Mean (mn) and standard deviations (sd) by month for forage ratio (FRATIO), prey abundance (PABUND), prey richness (RICH) and prey diversity (DINDEX) statistics for a) fry and b) smolts.*__
```{r MetricsDataAll, warning=FALSE, message=FALSE, echo=FALSE}

# For the chinook project, ran all the code to produce stuff here. 
# This time, we will only be referring to objects already created by the main script.

```


### Metrics by month across transects

__*Table 2a, b: Mean (mn) and standard deviations (sd) by transects within months for forage ratio (FRATIO), prey abundance (PABUND), prey richness (RICH) and prey diversity (DINDEX) statistics for a) fry and b) smolts.*__
```{r MetricsDataGroupsMoandTrans, warning=FALSE, message=FALSE, echo=FALSE}

# some other table making code ... 

```
Text goes here 

```{r Figure_4, warning=FALSE, message=FALSE, echo=FALSE, fig.width=5, fig.height=7, dpi=300, fig.align="center"}

# Below is a figure caption
```
__*Figure 4: Metrics for each group by MONTH, faceted by TRANSECT.*__


## Heading 2
More text

### Heading 3

## R packages

dplyr: A fast, consistent set of tools for working with data frame like objects, both in memory and out of memory.  

ggplot2: A system for 'declaratively' creating graphics based on ``The Grammar of Graphics''. You provide the data, tell 'ggplot2' how to map variables to aesthetics, what graphical primitives to use, and it takes care of the details. 
Hmisc: Contains many functions useful for data analysis, high-level graphics, utility operations, functions for computing sample size and power, simulation, importing and annotating datasets,
imputing missing values, advanced table making, variable clustering, character string manipulation, conversion of R objects to LaTeX and html code, and recoding variables. Used here to add standard deviation to as lines to ggplot.

knitr: Provides a general-purpose tool for dynamic report generation in R using Literate Programming techniques. Used here for Markdown formatting as html or pdf.

lubridate : Functions to work with date-times and time-spans: fast and user friendly parsing of date-time data, extraction and updating of components of a date-time (years, months, days, hours, minutes, and seconds), algebraic manipulation on date-time and time-span objects. Used here to pool dates to month.

markdown: R Markdown allows the use of knitr and Pandoc in the R environment. This package translates R Markdown to standard Markdown that knitr and Pandoc render to the desired output format (e.g., PDF, HTML, Word, etc).

purrr: A complete and consistent functional programming toolkit for data manupulation in R.  

readxl: Imports excel files into R.  

reshape2: Flexibly restructure and aggregate data using just two functions: melt and 'dcast'. Used here for ggplot() support. 

tibble: Functions for manipulating the tibble data format in R. Used here for the deframe() function.  

vegan: Ordination methods, diversity analysis and other functions for community and vegetation ecologists.  

